---
title: "Business Dynamics Database (BDS)"
format:
  html:
    toc: true
    toc-location: right-body
    code-fold: true
    code-overflow: wrap
    embed-resources: true
    df-print: kable
execute:
  cache: true
---



```{r}
#| echo: false
#| output: false
library(broom)
library(tidyverse)
library(arrow)
library(DT)


COLLECTION <- "bds"

```

# Overview

> BDS provides annual measures of business dynamics (such as job creation and destruction, establishment births and deaths, and firm startups and shutdowns) for the economy overall and aggregated by establishment and firm characteristics.

The revision available in `pubdata` is 2023.
Some 2022 tables are also included and should be available as long as source files exist at original URLs.

Data source links.

```{r}
#| echo: false
#| results: asis
pubdata::meta(COLLECTION, print = FALSE)$urls %>%
  iwalk(\(url, name) {
    cat(str_glue("  - [{name}]({url})"), "\n")
  })

```


# Tables

List of all tables in the collection.

```{r}
#| echo: false
#| column: page
# eval: false

descriptions <- pubdata::ls(COLLECTION, detail = TRUE) %>%
  select(key, description) %>%
  filter(!str_starts(key, "raw"))

dimensions <- descriptions$key %>%
  map(\(key) {
    df <- pubdata::path(COLLECTION, key) %>%
      open_dataset()
    list(key = key, rows = nrow(df), cols = ncol(df))
  }) %>%
  bind_rows()

inner_join(descriptions, dimensions, by = "key") %>%
  datatable()

```


# Columns

Listing of all columns in each table.

```{r}
#| echo: false
#| column: page
# eval: false

pubdata::ls(COLLECTION) %>%
  grepv("^raw_", ., invert = TRUE) %>%
  set_names() %>%
  map(\(key) {
    schema <- pubdata::meta(COLLECTION, key, print = FALSE)$schema %>%
      bind_rows(.id = "column") %>%
      relocate(column, type, desc)
    stats <- pubdata::get(COLLECTION, key) %>%
      map(\(values) list(na = sum(is.na(values)), unique = length(unique(values)))) %>%
      bind_rows(.id = "column")
    inner_join(schema, stats, by = "column")
  }) %>%
  bind_rows(.id = "key") %>%
  datatable(filter = "top")
```




# Examples

Author: Diveesha Vanchireddy

## Metro vs non-metro exit rates

How do the exit rates of establishments differ pre and post covid in metro vs non metros?

```{r}
#| warning: false

# load the processed metro/nonmetro dataset
df <- pubdata::get("bds", "2022_met") %>%
  mutate(
    year = as.integer(year),
    estabs_exit_rate = as.numeric(estabs_exit_rate),
    metro_status = case_when(
      metro == "M" ~ "metro",
      metro == "N" ~ "nonmetro",
      TRUE ~ NA_character_
    ),
    covid_period = ifelse(year < 2020, "pre_covid", "post_covid")
  ) %>%
  filter(year >= 2016 & year <= 2022, !is.na(metro_status))

# summarise
summary_df <- df %>%
  group_by(year, metro_status) %>%
  summarise(avg_exit_rate = mean(estabs_exit_rate, na.rm = TRUE), .groups = "drop")

# plot
ggplot(summary_df, aes(x = factor(year), y = avg_exit_rate, fill = metro_status)) +
  geom_col(position = "dodge") +
  geom_vline(
    xintercept = which(sort(unique(summary_df$year)) == 2020),
    linetype = "dashed", color = "red", size = 1
  ) +
  labs(
    title = "Exit Rate of Establishments (Metro vs Nonmetro, 2016–2022)",
    y = "Average Exit Rate",
    x = "Year"
  ) +
  theme_minimal()

```

The data indicate that metro establishments were more vulnerable to
pandemic-related shocks, hence we see a larger establishment exit rate.
Nonmetro establishments showed a smaller relative increase.This suggests
that rural economies may have been somewhat more protected from pandemic
disruptions which is an important factor to consider when working with
business dynamics in rural areas.

## Establishment exit and job creation

How does the job creation rate differ with establishment exit rate?

```{r}

# Load and preprocess correctly
df <- pubdata::get("bds", "2022_met") %>%
  mutate(
    job_creation_rate = as.numeric(job_creation_rate),
    establishment_exit_rate = as.numeric(estabs_exit_rate),
    metro_status = case_when(
      metro == "M" ~ "metro",
      metro == "N" ~ "nonmetro",
      TRUE ~ NA_character_
    ),
  ) %>%
  filter(
    metro_status %in% c("metro", "nonmetro"),  # keep only metro and nonmetro
    !is.na(job_creation_rate),
    !is.na(establishment_exit_rate)
  )

# Summarize the relationship (linear regression by metro status)
relationship_summary <- df %>%
  group_by(metro_status) %>%
  do(tidy(lm(establishment_exit_rate ~ job_creation_rate, data = .))) %>%
  filter(term == "job_creation_rate") %>%
  select(metro_status, estimate, std.error, p.value)

print(relationship_summary)

# Scatterplot with regression lines
ggplot(df, aes(x = job_creation_rate, y = establishment_exit_rate, color = metro_status)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Job Creation Rate vs Establishment Exit Rate",
    subtitle = "Separate regression lines for Metro (Urban) and Nonmetro (Rural) Areas",
    x = "Job Creation Rate",
    y = "Establishment Exit Rate",
    caption = "Printed table shows slope estimates by area type"
  ) +
  theme_minimal()

```

The linear model above shows that in there is positive relationship
between job creation rate and the establishment exit rate in both metro
areas and non metro areas but the slope of non metro areas is higher
than metro areas which signifies that there is higher instability in
rural areas compared to urban areas.

## Jobs in metro vs rural areas

What is the share of total employment in rural vs metro areas?

```{r}
df_met <- pubdata::get("bds", "2022_met") %>%
  mutate(
    emp = as.numeric(emp),  # total employment
    metro_status = case_when(
      metro == "M" ~ "metro",
      metro == "N" ~ "nonmetro",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(metro_status %in% c("metro","nonmetro"))

# Summed employment by type
emp_summary <- df_met %>%
  group_by(metro_status) %>%
  summarise(total_emp = sum(emp, na.rm = TRUE), .groups = "drop")

# Add share
total_us_emp <- sum(emp_summary$total_emp, na.rm = TRUE)

emp_summary %>%
  mutate(share = total_emp / total_us_emp)

```

We can see that most of the jobs are concentrated in the metro areas
with the employment share being 96% and the non metro's share of
employment being 3%

## Metro vs rural growth

Which area is growing more(firms,employment) in metro vs non metro areas?
This also accounts for the change before and after the COVID pandemic.

```{r}
# Load data (replace year key if needed)
df <- pubdata::get("bds", "2022_met") %>%
  mutate(
    metro_status = case_when(
      metro == "M" ~ "metro",
      metro == "N" ~ "nonmetro",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(metro_status))

# Summarize total firms and employment by metro status and year
summary_df <- df %>%
  group_by(year, metro_status) %>%
  summarise(
    total_firms = sum(firms, na.rm = TRUE),
    total_emp   = sum(emp, na.rm = TRUE),
    .groups = "drop"
  )

# Plot trends
p1 <- ggplot(summary_df, aes(x = year, y = total_firms, color = metro_status)) +
  geom_line(size = 1.2) +
  labs(title = "Firm Growth: Metro vs Nonmetro", y = "Number of Firms", x = "Year") +
  theme_minimal()

p2 <- ggplot(summary_df, aes(x = year, y = total_emp, color = metro_status)) +
  geom_line(size = 1.2) +
  labs(title = "Employment Growth: Metro vs Nonmetro", y = "Total Employment", x = "Year") +
  theme_minimal() + expand_limits(y=0)

# ---- Growth rates (indexed to first year) ----
summary_growth <- summary_df %>%
  group_by(metro_status) %>%
  arrange(year) %>%
  mutate(
    firms_index = total_firms / first(total_firms) * 100,
    emp_index   = total_emp / first(total_emp) * 100
  ) %>%
  ungroup()

p3 <- ggplot(summary_growth, aes(x = year, y = firms_index, color = metro_status)) +
  geom_line(size = 1.2) +
  labs(title = "Firm Growth Rate: Indexed to First Year", y = "Index (First Year = 100)", x = "Year") +
  theme_minimal() +
  expand_limits(y = 0)

p4 <- ggplot(summary_growth, aes(x = year, y = emp_index, color = metro_status)) +
  geom_line(size = 1.2) +
  labs(title = "Employment Growth Rate: Indexed to First Year", y = "Index (First Year = 100)", x = "Year") +
  theme_minimal() +
  expand_limits(y = 0)


print(p1)
print(p2)
print(p3)
print(p4)


# Compute growth rates pre vs post COVID
growth_summary <- summary_df %>%
  group_by(metro_status) %>%
  summarise(
    firms_growth = (last(total_firms) - first(total_firms)) / first(total_firms) * 100,
    emp_growth   = (last(total_emp) - first(total_emp)) / first(total_emp) * 100,
    .groups = "drop"
  )

print(growth_summary)

```

We can see that in metro areas the firms grew by 60% post pandemic and
only 2.5% where as for the employment rate in non metro areas there has
been a 92% growth and only a 32% growth in non metro areas. However
according to the plots we cannot clearly comprehend the growth for the
non metro regions so we index to first year(base year). By doing this we
can see which area is growing faster proportionally.Non metro regions
have a lesser growth rate compared to metro regions(We can see this more
clearly after we indexed it with a base of 100).

## Multi-establishment firms

Do rural areas have more single establishment firms vs multi-establishment firms?

```{r}
df <- pubdata::get("bds", "2022_met") %>%
  mutate(
    metro_status = case_when(
      metro == "M" ~ "metro",
      metro == "N" ~ "nonmetro",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(metro_status))

# Calculate single vs multi establishment counts
df_summary <- df %>%
  group_by(year, metro_status) %>%
  summarise(
    total_firms = sum(firms, na.rm = TRUE),
    total_estabs = sum(estabs, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    single_estab_firms = total_firms,                     # every firm has at least one estab
    multi_estab_extra  = total_estabs - total_firms,      # establishments beyond 1 per firm
    share_single_estab = total_firms / total_estabs * 100 # % of establishments that are single-unit
  )

print(df_summary)

# Visualization: share of single-unit establishments
ggplot(df_summary, aes(x = year, y = share_single_estab, color = metro_status)) +
  geom_line(size = 1.2) +
  labs(
    title = "Share of Single-Establishment Firms",
    y = "Percent Single-Establishment",
    x = "Year"
  ) +
  theme_minimal()


```

Rural areas continue to depend more on small, independent firms, while
metros are more dominated by multi-location businesses. But both areas
are trending toward fewer single-establishment firms over time,
reflecting a national shift toward consolidation and chain expansion.

