---
title: "BDS development"
format:
  html:
    code-fold: true

---

```{r}
library(tidyverse)
library(yaml)
library(rvest)

logger::log_threshold(logger::DEBUG)
```


# Datasets

Descriptions and links to BDS source data files.

YAML metadata entry template:

```
raw_2023_fa:
  desc: BDS by Firm Age, 2023 revision
  type: raw
  path: raw/2023/bds2023_fa.csv
  url: https://www2.census.gov/programs-surveys/bds/tables/time-series/2023/bds2023_fa.csv
```


```{r}
dataset_list_page <- "tmp/bds_2023_datasets.html"
# download_file("https://www.census.gov/programs-surveys/bds/data.Datasets.html", dataset_list_page)

page <- read_html(dataset_list_page)

nodes <- page |> html_elements("a")

links <- tibble(
    url = html_attr(nodes, "href"),
    desc = html_text2(nodes)
  ) |>
  filter(str_detect(url, "\\.csv(\\?.*)?$")) |>
  mutate(
    desc = desc |>
      str_squish() |>
      str_remove("\\s*\\[.*?\\]\\s*") |>
      str_c("BDS ", ... = _, ", 2023 revision"),
    file = url |> basename() |> str_remove("\\?.*$"),
    key = str_remove(file, "\\.csv$") |> str_remove("bds") |> str_to_lower(),
    path = paste0("raw/2023/", file)
  ) |>
  distinct(url, .keep_all = TRUE)

yaml_list <- links |>
  mutate(
    entry = pmap(
      list(key, desc, path, url),
      \(key, desc, path, url) {
        setNames(
          list(
            list(
              desc = desc,
              type = "raw",
              path = path,
              url = url
            )
          ),
          paste0("raw_", key)
        )
      }
    )
  ) |>
  pull(entry) |>
  reduce(c)

yaml_str <- as.yaml(yaml_list)

cat(yaml_str)

```



```{r}
pubdata::ls("bds", "2023.*cty", TRUE)


```

# Schema

Parse codebook PDF to use in YAML schema constructon.

```{r}
# download_file("https://www.census.gov/content/dam/Census/programs-surveys/business-dynamics-statistics/codebook-glossary.pdf", "tmp/codebook-glossary.pdf")

# library(pdftools)
# txt <- pdf_text("tmp/codebook-glossary.pdf")[1:6]
# writeLines(txt, "tmp/codebook-glossary-pages1-6.txt")


txt <- readLines("tmp/codebook-glossary-pages1-6.txt")

# Remove page number lines (standalone numbers or lines with only spaces and a number)
txt <- txt[!str_detect(txt, "^\\s*\\d+\\s*$")]
# Remove the header line
txt <- txt[!str_detect(txt, "BDS CODEBOOK AND GLOSSARY")]
txt <- txt[!str_detect(txt, "BDS VARIABLE DEFINITIONS AND CODES")]

# Collapse into single string for regex parsing
full <- paste(txt, collapse = "\n")

# Split on variable names (left-aligned words at start of line)
# Variable names appear as lines starting with no leading whitespace
chunks <- str_split(full, "(?=\n[a-z][a-z_0-9]*\n)")[[1]]
# Drop preamble
chunks <- chunks[str_detect(chunks, "Values:|Description:")]

parse_chunk <- function(chunk) {
  # Extract variable name (first non-blank line with no indentation)
  name <- str_extract(chunk, "(?<=\n)[a-z][a-z_0-9]*(?=\n)")
  if (is.na(name)) return(NULL)

  # Extract values
  values_match <- str_match(chunk, "Values:\\s*(.*?)(?=\\n\\s*\\n|\\n\\s*Description:)")
  # values <- str_squish(values_match[, 2]) %|% ""
  values <- str_squish(values_match[, 2]) |> replace_na("")
  
  # Extract description (everything after "Description:" until end)
  desc_match <- str_match(chunk, regex("Description:\\s*(.*)", dotall = TRUE))
  desc_raw <- desc_match[, 2]
  if (is.na(desc_raw)) return(NULL)

  # Clean up description: collapse multi-line, normalize whitespace
  desc_lines <- str_split(desc_raw, "\n")[[1]]
  desc_lines <- str_squish(desc_lines)
  desc_lines <- desc_lines[desc_lines != ""]
  desc <- paste(desc_lines, collapse = "\n")

  # Build desc block: description + values if present
  desc_block <- desc
  if (!is.na(values) && values != "") {
    desc_block <- paste0(desc, "\nValues: ", values)
  }

  list(name = name, desc_block = desc_block)
}

parsed <- map(chunks, parse_chunk) |> compact()

# Build YAML list
yaml_entries <- map(parsed, \(p) {
  setNames(list(list(type = "character", desc = p$desc_block)), p$name)
}) |> reduce(c)

yaml_out <- as.yaml(yaml_entries, indent.mapping.sequence = TRUE)

writeLines(yaml_out, "tmp/codebook.yml")
```


Schema stubs from raw.


```{r}
dblq <- \(x) paste0("\"", x, "\"")
yaml_text <- ls("bds", "raw.*2023") %>%
  str_remove("^raw_") %>%
  set_names() %>%
  map(function(key) {
    raw_key <- paste0("raw_", key)
    met <- meta("bds", raw_key, FALSE)
    mask <- str_replace(key, "2023", "{revision}")
    sch <- str_remove(key, ".*2023_") |>
      str_split_1("_") |>
      c("year", ... = _, "core") |>
      paste0("*schema_", ... = _) |>
      paste(collapse = ", ") |>
      paste0("[", ... = _, "]")
    list(
      mask = dblq(mask),
      desc = str_replace(met$desc, "2023", "{revision}") |> dblq(),
      type = "table",
      path = paste0("proc/{revision}/bds", mask, ".pq") |> dblq(),
      depends = str_replace(raw_key, "2023", "{revision}") |> dblq(),
      schema = list(
        "<<" = sch
      )
    )
  }) %>%
    as.yaml() %>%
    str_replace_all("'", "")

cat(yaml_text)
clipr::write_clip(capture.output(cat(yaml_text)))

```



```{r}
get("bds", "raw_2023_fac_fzc")
```

Compare alignment of raw csv columns and schema columns

```{r}
key <- "2023_sec"
count_cols <- c("sector")
raw_head <- get("bds", paste0("raw_", key)) %>%
  read_csv(n_max = 1, show_col_types = FALSE)
meta_schema <- meta("bds", key, FALSE)$schema
non_flag_cols <- names(meta_schema) %>% grepv("_f$", ., invert = TRUE)
meta_schema <- meta_schema[non_flag_cols]
# compare overlap
cmp <- full_join(
  tibble(col = names(raw_head), raw = "raw"),
  tibble(col = names(meta_schema), meta = "meta"),
  by = "col"
) %>%
  mutate(res = paste(raw, meta, sep = "+"))

cmp %>% count(res)
cmp %>% filter(res != "raw+meta")

# compare ordering
cmp1 <- bind_cols(list(raw = names(raw_head), meta = names(meta_schema)))
cmp1 %>% filter(raw != meta)

x <- get("bds", paste0("raw_", key)) %>%
  read_csv(col_types = list(.default = "c"), show_col_types = FALSE)
x %>% count(across(all_of(count_cols)))
```

bulk test

```{r}
schema_alignment <- ls("bds", "raw.*2022") %>%
  map(function(raw_key) {
    key <- str_remove(raw_key, "^raw_")
    raw_head <- get("bds", raw_key) %>%
      read_csv(n_max = 1, show_col_types = FALSE)
    meta_schema <- meta("bds", key, FALSE)$schema
    non_flag_cols <- names(meta_schema) %>% grepv("_f$", ., invert = TRUE)
    meta_schema <- meta_schema[non_flag_cols]
    # compare overlap
    overlap <- full_join(
      tibble(col = names(raw_head), raw = "raw"),
      tibble(col = names(meta_schema), meta = "meta"),
      by = "col"
    ) %>%
      mutate(res = paste(raw, meta, sep = "+")) %>% 
      filter(res != "raw+meta") %>% 
      nrow()

    # compare order
    order <- bind_cols(list(raw = names(raw_head), meta = names(meta_schema))) %>%
      filter(raw != meta) %>%
      nrow()
    list(key = key, raw_key = raw_key, nraw = ncol(raw_head), nsch = length(meta_schema), overlap = overlap, order = order)
  }) %>%
    bind_rows()

schema_alignment %>%
  filter(overlap > 0 | order > 0)


```



```{r}
ls("bds", "2023.*ez")
meta("bds", "2023_sec_ez")
meta("bds", "2023_sec_ez", FALSE)$schema %>%
  map(\(x) x$type) %>%
  {tibble(column = names(.), type = unlist(.))} %>%
  count(type)
```

## handle NaN values

[Suppression flags](https://www.census.gov/programs-surveys/bds/documentation/methodology.html)

> In addition to noise infusion, cells with fewer than three firms are suppressed with a publication flag 'D'. In addition, cells with identified data quality concerns are suppressed with a publication flag 'S'. Cells that are "structurally missing" or "structurally zero" are indicated with a publication flag 'X'. Finally, rate cells that cannot be calculated are indicated with a publication flag 'N'.


```{r}
x <- get("bds", "raw_2023_sec_ez") %>%
  read_csv(col_types = list(.default = "c"))
x %>%
  mutate(across(everything(), str_squish)) %>%
  mutate(across(everything(), \(x) if_else(str_detect(x, "^-?\\d+(\\.\\d+)?$"), "num", x))) %>%
  map(table)

```


# Build all


```{r}
all_key_dims <- ls("bds") %>%
  grepv("^raw_", ., invert = TRUE) %>%
  set_names() %>%
  map(\(k) get("bds", k) %>% dim() %>% set_names(c("nrow", "ncol"))) %>%
  bind_rows(.id = "key")

all_key_dims
```