# BDS Collection Analysis Template
# Author: Diveesha
# Purpose: Document and explore the BDS collection datasets
## 🗂️ BDS Collection Overview

The **Business Dynamics Statistics (BDS)** is a dataset published by the U.S. Census Bureau that provides annual measures of business dynamics (e.g., job creation, job destruction, establishment openings/closings) in the U.S. economy, disaggregated by firm age, size, industry, and geography.

BDS is compiled using the **Longitudinal Business Database (LBD)**, covering all employer businesses in the U.S. that have a valid EIN.

🔗 Official documentation:  
- [BDS Homepage](https://www.ers.usda.gov/data-products/rural-urban-continuum-codes)  
- [Data Documentation](https://www.census.gov/programs-surveys/bds/technical-documentation/methodology.html)

### 📦 Collection Structure in `pubdata`

In the `pubdata` package, the BDS collection includes processed and raw datasets grouped by geography and revision year. Examples include:

| Dataset Name       | Type    | Description                                       |
|--------------------|---------|---------------------------------------------------|
| `bds_2022_met`     | table   | BDS by metro/non-metro areas, 2022 revision      |
| `raw_bds_2022_met` | raw     | Raw version of the above table                   |
| `bds_2022_naics`   | table   | BDS by NAICS industry classification, 2022       |
| `bds_2021_size`    | table   | BDS by firm size, 2021 revision                  |

The goal of this notebook is to:
- Document each dataset with structure, size, and contents
- Provide variable-level metadata (codebooks)
- Generate summary statistics
- Explore usage and reproducibility templates for future research






```{r}
library(pubdata)
library(readr)
library(purrr)

# Step 1: Get all dataset keys in the raw_2022_met collection
dataset_keys <- pubdata::ls("bds", "raw_2022_met")

# Step 2: Use pubdata::get() to get file paths, then read with read_csv
df_list <- dataset_keys %>%
  set_names() %>%
  map(~ {
    csv_path <- pubdata::get("bds", .x)  # assuming this returns a file path
    read_csv(csv_path)
  })

# ✅ Check what's inside
names(df_list)
df_list[["raw_2022_met"]]


```



```{r}
descriptions <- dataset_keys %>%
  set_names() %>%
  map_chr(~ pubdata::meta("bds", .x)$desc)
  
desc_df <- tibble(
  key = names(descriptions),
  description = descriptions
)
desc_df
```


# Dataset summary
```{r}
dataset_summary <- map_dfr(names(df_list), function(name) {
  df <- df_list[[name]]
  tibble(
    Key = name,
    Type = ifelse(grepl("^raw", name), "raw", "table"),
    Description = desc_df$description,
    Num_Variables = ifelse(is.null(dim(df)), NA, ncol(df)),
    Num_Observations = ifelse(is.null(dim(df)), NA, nrow(df))
  )
})
print("Dataset Summary:")
print(dataset_summary)
```

# Generate codebook for a dataset

```{r}
generate_codebook <- function(df) {
  tibble(
    Variable = names(df),
    Type = map_chr(df, ~ class(.x)[1]),
    Description =  desc_df$description,
  )
}
codebook_2022_met <- generate_codebook(dataset_keys)
print("Codebook for bds_2022_met:")
print(codebook_2022_met)
```




```{r}
install.packages("collapsibleTree")
```
```{r}
df_list
```
```{r}
make_variable_skeleton <- function(df_list, file_path) {
  df <- read_csv(file_path, show_col_types = FALSE)
  tibble(
    dataset_key = dataset_key,
    variable = names(df),
    description = NA_character_
  )
}
```

```{r}
library(dplyr)
library(purrr)
library(collapsibleTree)

generate_codebook <- function(df, desc_df, dataset_name) {
  tibble(
    Dataset = dataset_name,
    Variable = names(df),
    Description = map_chr(names(df), function(var) {
      matched <- desc_df %>% filter(key == var)
      if (nrow(matched) > 0) matched$description else "No description available"
    })
  )
}

codebook_bds_2022 <- generate_codebook(desc_df, desc_df$description, "raw_2022_met")

```

# Basic summary statistics for numeric columns

```{r}
basic_stats <- function(df) {
  df %>%
    summarise(across(where(is.numeric), list(
      mean = ~mean(.x, na.rm = TRUE),
      sd = ~sd(.x, na.rm = TRUE),
      min = ~min(.x, na.rm = TRUE),
      max = ~max(.x, na.rm = TRUE)
    ), .names = "{.col}_{.fn}")) %>%
    pivot_longer(cols = everything(), names_to = "stat", values_to = "value")
}
summary_stats_2022_met <- basic_stats(pubdata::bds_2022_met)
print("Summary Stats for bds_2022_met:")
print(summary_stats_2022_met)

```
