---
title: "BEA Regional Economic Accounts"
format:
  html:
    toc: true
    toc-location: right-body
    code-fold: true
    code-overflow: wrap
    embed-resources: true
    df-print: kable
execute:
  cache: true
---


```{r}
#| echo: false
#| output: false
library(broom)
library(tidyverse)
library(arrow)
library(DT)

COLLECTION <- "bea_reg"

```

# Overview

> The regional economic accounts tell us about the geographic distribution of U.S. economic activity and growth. The estimates of gross domestic product by state and state and local area personal income, and the accompanying detail, provide a consistent framework for analyzing and comparing individual state and local area economies.

Data source links.

```{r}
#| echo: false
#| results: asis
pubdata::meta(COLLECTION, print = FALSE)$urls %>%
  iwalk(\(url, name) {
    cat(str_glue("  - [{name}]({url})"), "\n")
  })

```


# Tables

List of all tables in the collection.

```{r}
#| echo: false
#| column: page
# eval: false

descriptions <- pubdata::ls(COLLECTION, detail = TRUE) %>%
  select(key, description) %>%
  filter(!str_starts(key, "raw"))

dimensions <- descriptions$key %>%
  map(\(key) {
    pq_path <- pubdata::path(COLLECTION, key)
    if (!file.exists(pq_path)) pubdata::get(COLLECTION, key)
    df <- open_dataset(pq_path)
    list(key = key, rows = nrow(df), cols = ncol(df))
  }) %>%
  bind_rows()

inner_join(descriptions, dimensions, by = "key") %>%
  datatable()

```


# Columns

Listing of all columns in each table.

```{r}
#| echo: false
#| column: page
# eval: false

pubdata::ls(COLLECTION) %>%
  grepv("^raw_", ., invert = TRUE) %>%
  set_names() %>%
  map(\(key) {
    schema <- pubdata::meta(COLLECTION, key, print = FALSE)$schema %>%
      bind_rows(.id = "column") %>%
      relocate(column, type, desc)
    stats <- pubdata::get(COLLECTION, key) %>%
      map(\(values) list(na = sum(is.na(values)), unique = length(unique(values)))) %>%
      bind_rows(.id = "column")
    inner_join(schema, stats, by = "column")
  }) %>%
  bind_rows(.id = "key") %>%
  datatable(filter = "top")
```



# Examples

Coming soon.
