---
title: "County Business Patterns (CBP)"
format:
  html:
    toc: true
    toc-location: right-body
    code-fold: true
    code-overflow: wrap
    embed-resources: true
    df-print: kable
execute:
  cache: true
---



```{r}
#| echo: false
#| output: false
library(broom)
library(tidyverse)
library(arrow)
library(DT)

COLLECTION <- "cbp"

```

# Overview

> County Business Patterns (CBP) is an annual series that provides subnational economic data for establishments with paid employees by industry and employment size. This series includes the number of establishments, employment during the week of March 12, first quarter payroll, and annual payroll.

Data source links.

```{r}
#| echo: false
#| results: asis
pubdata::meta(COLLECTION, print = FALSE)$urls %>%
  iwalk(\(url, name) {
    cat(str_glue("  - [{name}]({url})"), "\n")
  })

```


# Tables

List of all tables in the collection.

```{r}
#| echo: false
#| column: page
# eval: false

descriptions <- pubdata::ls(COLLECTION, detail = TRUE) %>%
  select(key, description) %>%
  filter(!str_starts(key, "raw"))

dimensions <- descriptions$key %>%
  map(\(key) {
    pq_path <- pubdata::path(COLLECTION, key)
    if (!file.exists(pq_path)) pubdata::get(COLLECTION, key)
    df <- open_dataset(pq_path)
    list(key = key, rows = nrow(df), cols = ncol(df))
  }) %>%
  bind_rows()

inner_join(descriptions, dimensions, by = "key") %>%
  datatable()

```


# Columns

Listing of all columns in each table.

```{r}
#| echo: false
#| column: page
# eval: false

pubdata::ls(COLLECTION) %>%
  grepv("^raw_", ., invert = TRUE) %>%
  set_names() %>%
  map(\(key) {
    schema <- pubdata::meta(COLLECTION, key, print = FALSE)$schema %>%
      bind_rows(.id = "column") %>%
      relocate(column, type, desc)
    stats <- pubdata::get(COLLECTION, key) %>%
      map(\(values) list(na = sum(is.na(values)), unique = length(unique(values)))) %>%
      bind_rows(.id = "column")
    inner_join(schema, stats, by = "column")
  }) %>%
  bind_rows(.id = "key") %>%
  datatable(filter = "top")
```




# Examples

## Dane county sectors in 2012


```{r}
pubdata::path("cbp", "2012_co") %>%
  open_dataset() %>%
  filter(fipstate == "55", fipscty == "025") %>%
  filter(naics != "------", str_ends(naics, "----")) %>%
  mutate(naics = str_sub(naics, 1, 2)) %>%
  select(naics, est, emp, ap) %>%
  collect()

```
