---
title: "CHRR development"
format: html
---

Development notebook for CHRR

```{r}
library(tidyverse)
library(yaml)
library(readxl)
logger::log_threshold(logger::DEBUG)
```

# Downloadable files

The County Health Rankings & Roadmaps (CHRR) analytic dataset contains county-level measures for health outcomes, health behaviors, social and economic factors, and demographics, etc.

Start working with a single file (2024 analytic data)

```{r}
df <- read_csv("tmp/chrr/analytic_data2024.csv", show_col_types = FALSE, skip = 1)
df %>% head()
```

```{r}
ls("chrr")

meta("chrr", "raw_2024")
```


# Analytic data

## schema 

File layout and relevant measure/description are downloadable, using it to construct schema for YAML file (all CHRR analytic and trend data share the same schema in the official data dictionary).

# 2024
```{r}
#download_file("https://www.countyhealthrankings.org/sites/default/files/media/document/DataDictionary_2024.xlsx", "tmp/chrr/DataDictionary_2024.xlsx")

df_layout_2024 <- read_excel("tmp/chrr/DataDictionary_2024.xlsx") |> filter(!is.na(`Variable Name`))
s <- list()
for (i in 1:nrow(df_layout_2024)) {
  r <- df_layout_2024[i, ]
  var <- r$`Variable Name`
  desc <- gsub("\n", " ", r$Measure)
  if (is.na(var) || var == "") next
  col_meta <- list(desc = desc, type = "character")
  if (grepl("flag", var, ignore.case = TRUE)) col_meta$type <- "integer"
  if (grepl("value|numerator|denominator|ci", var, ignore.case = TRUE)) col_meta$type <- "double"
  s[[var]] <- col_meta
}
cat(yaml::as.yaml(s))
```



# 2025
```{r}
#download_file("https://www.countyhealthrankings.org/sites/default/files/media/document/DataDictionary_2025.xlsx", "tmp/chrr/DataDictionary_2025.xlsx")

df_layout_2025 <- read_excel("tmp/chrr/DataDictionary_2025.xlsx") |> filter(!is.na(`Variable Name`))
s <- list()
for (i in 1:nrow(df_layout_2025)) {
  r <- df_layout_2025[i, ]
  var <- r$`Variable Name`
  desc <- gsub("\n", " ", r$Measure)
  if (is.na(var) || var == "") next
  col_meta <- list(desc = desc, type = "character")
  if (grepl("flag", var, ignore.case = TRUE)) col_meta$type <- "integer"
  if (grepl("value|numerator|denominator|ci", var, ignore.case = TRUE)) col_meta$type <- "double"
  s[[var]] <- col_meta
}
cat(yaml::as.yaml(s))
```

## main tables
```{r}
for (y in 2024:2025){
  x <- chrr_get(paste0(y)) 
  cat(y, dim(x), "\n")
}
```

Full datasets
```{r}
get("chrr", "2025")
```

Query from a parquet
```{r}
path("chrr", "2025") %>%
  arrow::open_dataset() %>%
  filter(state == "WI") %>%
  select(v001_rawvalue, v002_rawvalue) %>%
  collect()
```

Access all years as a single parquet dataset.
```{r}
path("chrr", "2025") %>%
  dirname() %>%
  arrow::open_dataset() %>%
  filter(state == "CA") %>%
  select(year, county, v138_cilow, v182_rawvalue, v131_other_data_1, v063_race_asian) %>%
  collect()
```


## Variable-year table

[2023 Data Dictionary](https://www.countyhealthrankings.org/sites/default/files/media/document/2023%20Data%20Dictionary%20(XLSX).xlsx) uses a slightly different filename pattern than other years.
Manually renamed file to align with other years.

```{r}
# 2023 Data Dictionary uses a slightly different file pattern than other years
# download_file(https://www.countyhealthrankings.org/sites/default/files/media/document/2023%20Data%20Dictionary%20(XLSX).xlsx, "tmp/chrr/DataDictionary_2023.xlsx")


# find data dictionary files and extract year
files <- list.files("tmp/chrr", pattern = "DataDictionary.*\\.xlsx$", full.names = TRUE)
f_yr_extract <- tibble(path = files, year = str_extract(files, "\\d{4}")) |> filter(!is.na(year))
f_yr_extract

# read variables
vars_by_year <- list()
for (i in 1:nrow(f_yr_extract)) {
  year <- f_yr_extract$year[i]
  path <- f_yr_extract$path[i]
  df_year <- read_excel(path)
  var_names <- df_year$`Variable Name` |> na.omit() |> unique()
  vars_by_year[[year]] <- var_names
}

# create list and build presence matrix
variables <- unlist(vars_by_year) |> sort() |> unique()
var_change_df <- data.frame(variable = variables)
for (yr in names(vars_by_year)) { 
  var_change_df[[yr]] <- var_change_df$variable %in% vars_by_year[[yr]]
}

head(var_change_df, 10)
write_csv(var_change_df, "tmp/chrr/variable_year.csv")
```

# Trend data

## schema

[2024 CHR Trends Data Documentation](https://www.countyhealthrankings.org/sites/default/files/media/document/Trends%20documentation%202024.pdf) Unlike the analytic datasets (wide format), trend data are in long format, where each row corresponds to a measure and period combination. 
The schema has remained consistent across releases since 2021.

Numeric columns are read as doubles or integers where applicable.

```{r}
# download_file("https://www.countyhealthrankings.org/sites/default/files/media/document/Trends%20documentation%202024.pdf","tmp/chrr/Trends_documentation_2024.pdf")

df <- read_csv("tmp/chrr/chr_trends_csv_2024.csv", show_col_types = FALSE)
s <- list()
for (colname in names(df)) {
  s1 <- list(
    desc = "",
    type = "character"
  )

  # numeric types
  if (colname %in% c("numerator", "denominator", "rawvalue", "cilow", "cihigh"))
    s1$type <- "double"
  if (colname %in% c("measureid", "chrreleaseyear", "differflag", "trendbreak"))
    s1$type <- "integer"

  # column descriptions
  if (colname %in% c("yearspan")) s1$desc <- "The years of data this estimate represents"
  if (colname %in% c("measurename")) s1$desc <- "The name of the measure"
  if (colname %in% c("statecode")) s1$desc <- "The two-digit FIPS code for the state"
  if (colname %in% c("countycode")) s1$desc <- "The three-digit FIPS code for the county"
  if (colname %in% c("county")) s1$desc <- "The name of the county"
  if (colname %in% c("state")) s1$desc <- "The two-letter abbreviation for the state"
  if (colname %in% c("numerator")) s1$desc <- "The measure numerator"
  if (colname %in% c("denominator")) s1$desc <- "The measure denominator"
  if (colname %in% c("rawvalue")) s1$desc <- "The value of the measure"
  if (colname %in% c("cilow")) s1$desc <- "The lower bound of the confidence interval"
  if (colname %in% c("cihigh")) s1$desc <- "The upper bound of the confidence interval"
  if (colname %in% c("measureid")) s1$desc <- "The County Health Rankings measure identifier"
  if (colname %in% c("chrreleaseyear")) s1$desc <- "The County Health Rankings release year the data was included in"
  if (colname %in% c("differflag")) s1$desc <- "A flag to indicate that the estimate included in this release differs from the data included in the corresponding Annual Data Release."
  if (colname %in% c("trendbreak")) s1$desc <- "A flag to indicate that the year of the estimate represents the start of a new trend. The data from that point forward should not be used with prior years of data to estimate trends."
  s[[colname]] <- s1
}

cat(yaml::as.yaml(s))
```




# Long format

All non-identifier variables start with `vXXX`, followed by:

- one of "rawvalue", "numerator", "denominator", "cilow", "cihigh". Example: `v063_rawvalue`.
- `race_XXX`, by itself (for raw value) or followed by one of the above. Example: `v063_race_black`, `v063_race_black_cilow`.
- some variables have "other data". Example: `v136_other_data_2`.

```{r}
var_desc <- get("chrr", "raw_2025") %>%
  read_csv(n_max = 0, skip = 0) %>%
  names()

var_name <- get("chrr", "raw_2025") %>%
  read_csv(n_max = 0, skip = 1) %>%
  names()

# year and geographic identifiers do not start with "vXXX_"
var_name %>%
  grepv("^v\\d{3}_", ., invert = TRUE)

# all variables start with "vXXX_"
var_name %>%
  grepv("^v\\d{3}_", .)


# most variable match pattern (code, optional race, value), but not all:
x <- tibble(var = var_name, desc = var_desc) %>%
  separate_wider_regex(var, c(code = "v\\d{3}", race = ".*", "_", value = "[:alpha:]+"), too_few = "debug")
x

```


```{r}
x <- get("chrr", "raw_2025")

var_desc <- read_csv(x, n_max = 1, col_types = cols(.default = "c")) %>%
  pivot_longer(everything(), names_to = "description", values_to = "variable")

x1 <- read_csv(x, skip = 1, col_types = list(
  statecode = "c",
  countycode = "c",
  fipscode = "c",
  state = "c",
  county = "c",
  year = "i",
  county_clustered = "i",
  .default = "d")
) %>%
  pivot_longer(starts_with("v"), names_to = "variable") %>%
  left_join(var_desc, by = "variable")

x1
```


# column names in all years


```{r}
name_desc <- 2010:2025 %>%
  set_names() %>%
  map(\(y) {
    cat("----", y, "----", "\n")
    x <- get("chrr", paste0("raw_", y))
    x1 <- read_csv(x, n_max = 0, skip = 1, show_col_types = FALSE)
    x2 <- read_csv(x, n_max = 0, skip = 0, show_col_types = FALSE)
    list(var = names(x1), desc = names(x2))
  }) %>%
  bind_rows(.id = "year")


x2 <- name_desc %>%
  count(var)

# variable count by year
x2

# non-vXXX variables
x2 %>%
  filter(!str_detect(var, "^v\\d{3}"))


# changed in 2024 and 2025
name_desc %>%
  filter(str_detect(var, "^county_"))

```

Template for meta.yaml.

```{r}
name_desc %>%
  filter(year == 2023, !str_detect(var, "^v\\d{3}")) %>%
  pwalk(function(var, desc, ...) {
    type <- if (var == "year") "integer" else "character"
    cat(var, ":\n  desc: ", desc, "\n  type: ", type,  "\n", sep = "")
  })


```


## 2012 descriptions


Several duplicate column names:

```
   name  desc                                var             
   <chr> <chr>                               <chr>           
 1 X149  Access to healthy foods raw value   v030_rawvalue   
 2 X150  Access to healthy foods numerator   v030_numerator  
 3 X151  Access to healthy foods denominator v030_denominator
 4 X152  Access to healthy foods CI low      v030_cilow      
 5 X153  Access to healthy foods CI high     v030_cihigh     
 6 X221  Access to healthy foods raw value   v096_rawvalue   
 7 X222  Access to healthy foods numerator   v096_numerator  
 8 X223  Access to healthy foods denominator v096_denominator
 9 X224  Access to healthy foods CI low      v096_cilow      
10 X225  Access to healthy foods CI high     v096_cihigh   
```

```{r}
x <- get("chrr", "raw_2012") %>%
  read_csv(col_names = FALSE, n_max = 2) %>%
  add_column(col = c("desc", "var"), .before = 1) %>%
  pivot_longer(!col) %>%
  pivot_wider(id_cols = "name", names_from = "col")
x %>%
  filter(n() > 1, .by = desc)
dup_desc_vars <- x %>%
  filter(n() > 1, .by = desc) %>%
  pull(var)

```

Inspect values


```{r}
x1 <- get("chrr", "raw_2012") %>%
  read_csv(skip = 1)
x2 <- x1 %>%
  select(!starts_with("v"), all_of(dup_desc_vars))

```




# test all years


```{r}
devtools::load_all()
meta("chrr", "2011")

chrr_get("2021") %>%
  filter(str_detect(description, "uicide"))

```