---
title: "County Health Rankings & Roadmaps (CHRR)"
format:
  html:
    toc: true
    toc-location: right-body
    code-fold: true
    code-overflow: wrap
    embed-resources: true
    df-print: kable
execute:
  cache: true
---

```{r}
#| echo: false
#| output: false
library(tidyverse)
library(arrow)
library(DT)

COLLECTION <- "chrr"

```

# Overview

> County Health Rankings & Roadmaps (CHR&R), a program of the University of Wisconsin Population Health Institute, draws attention to why there are differences in health within and across communities. The program highlights policies and practices that can help everyone be as healthy as possible. CHR&R aims to grow a shared understanding of health, equity and the power of communities to improve health for all. This work is rooted in a long-term vision where all people and places have what they need to thrive. 

Data source links.

```{r}
#| echo: false
#| results: asis
pubdata::meta(COLLECTION, print = FALSE)$urls %>%
  iwalk(\(url, name) {
    cat(str_glue("  - [{name}]({url})"), "\n")
  })

```


# Tables

List of all tables in the collection.

```{r}
#| echo: false
#| column: page
# eval: false

descriptions <- pubdata::ls(COLLECTION, detail = TRUE) %>%
  select(key, description) %>%
  filter(!str_starts(key, "raw"))

dimensions <- descriptions$key %>%
  map(\(key) {
    df <- pubdata::get(COLLECTION, key)
    list(key = key, rows = nrow(df), cols = ncol(df))
  }) %>%
  bind_rows()

inner_join(descriptions, dimensions, by = "key") %>%
  datatable()

```


# Columns

Listing of all columns in each table.

```{r}
#| echo: false
#| column: page
# eval: false

pubdata::ls(COLLECTION) %>%
  grepv("^raw_", ., invert = TRUE) %>%
  set_names() %>%
  map(\(key) {
    schema <- pubdata::meta(COLLECTION, key, print = FALSE)$schema %>%
      bind_rows(.id = "column") %>%
      relocate(column, type, desc)
    stats <- pubdata::get(COLLECTION, key) %>%
      map(\(values) list(na = sum(is.na(values)), unique = length(unique(values)))) %>%
      bind_rows(.id = "column")
    inner_join(schema, stats, by = "column")
  }) %>%
  bind_rows(.id = "key") %>%
  datatable(filter = "top")
```


# Examples

## Variables across years

Variables, descriptions and variable parts.

```{r}
#| column: body-outset

all_year_vars <- pubdata::ls("chrr") %>%
  grepv("^raw_", ., invert = TRUE) %>%
  map(function(key) {
    pubdata::path("chrr", key) %>%
      open_dataset() %>%
      filter(fipscode == "00000") %>%
      select(year, variable, description) %>%
      collect() %>%
      separate_wider_regex(
        variable,
        cols_remove = FALSE,
        pattern = c(
          code = "^v\\d{3}",
          "_",
          race = "(?:race_[^_]+)?",
          "_?",
          value = ".*"
        )
      ) %>%
      relocate(year, variable, description)
  }) %>%
  bind_rows()

all_year_vars %>%
  datatable(filter = "top")

```


## Rural premature mortality

County relationship between percentage of rural population and premature age-adjusted mortality.

```{r}
#| warning: false
x <- pubdata::get("chrr", "2025")

x1 <- x %>%
  filter(countycode != "000") %>%
  filter(variable %in% c("v058_rawvalue", "v127_rawvalue")) %>%
  pivot_wider(id_cols = fipscode, names_from = variable) %>%
  rename(rural = v058_rawvalue, paam = v127_rawvalue)

x1 %>%
  ggplot(aes(rural, paam)) +
  geom_point(size = 0.5) +
  geom_smooth()

```


## National adult smoking over time

```{r}
pubdata::ls("chrr") %>%
  grepv("^raw_", ., invert = TRUE) %>%
  map_chr(\(key) pubdata::path("chrr", key)) %>%
  open_dataset() %>%
  filter(fipscode == "00000", variable == "v009_rawvalue") %>%
  select(year, value) %>%
  collect() %>%
  arrange(year)
 
```

