---
title: "Tidy Public Data"
format:
  html:
    date: today
    toc: true
    code-overflow: wrap
    embed-resources: true
    df-print: kable
---

"Like families, tidy datasets are all alike but every messy dataset is messy in its own way" ([Wickham, 2014][1]).

[1]: https://doi.org/10.18637/jss.v059.i10 "Wickham, H. (2014). Tidy Data. Journal of Statistical Software, 59(10), 1–23."

`pubdata` is an R package that provides a minimalist interface to a wide range of publicly available datasets in a standardized tidy table form. 

- Discover datasets by searching table and variable metadata.
- Programmatically retrieve datasets from their online source to local storage.
- YAML metadata format for easy entry of new datasets.
- Work with raw data files or tidy data frames.
- Standardized column names and embedded data types.
- Minimal modification of source data, all original data preserved.
- Parquet format for efficient storage and loading or out-of-core processing.
- Disk and memory cache for quick repeated access.

```{r}
#| echo: false
#| output: false
library(tidyverse)
```

# Datasets

Datasets are organized into collections by data provider and topic.

```{r}
#| echo: false
pubdata::ls() %>%
  map(\(col) {
    list(
      collection = col,
      description = pubdata::meta(col, print = FALSE)$desc,
      datasets = pubdata::ls(col) %>% grepv("^raw", ., invert = TRUE) %>% length()
    )
  }) %>%
  bind_rows()
```


# Package interface

It is recommended not to attach the package with `library(pubdata)`.
Instead, use the full namespace to call package functions, e.g. `pubdata::ls()`.
Package functions are a few and short, and using the full namespace helps avoid name conflicts.

Main package interface consists of these three functions:

- `ls()` lists available collections and datasets.
- `meta()` provides metadata information.
- `get()` retrieves a raw data file or a clean dataframe.


## `ls()`: list available datasets

Without parameters, returns a list of all available data collections.

```{r}
pubdata::ls()
```

With `detail = TRUE` returns collections with descriptions.

```{r}
pubdata::ls(detail = TRUE)
```

If `collection` is specified, returns the list of  available dataset keys.
For example, keys in the `"ers_rural"` collection:

```{r}
pubdata::ls(collection = "ers_rural")

# show descriptions and cache files
pubdata::ls(collection = "ers_rural", detail = TRUE) |>
  slice_sample(n = 6)
```

Use `pattern` parameter to list keys that match a regex pattern within the data collection.
For example, pattern `"^ruc_"` in the `"ers_rural"` collection will list all available Rural-Urban Continuum Code datasets from the USDA ERS rurality classifications collection.

```{r}
pubdata::ls(collection = "ers_rural", pattern = "^ruc_")
```

## `meta()`: show collection or dataset metadata

Obtain metadata of a collection or a specific dataset.

For *collections*, `meta()` will print:

  - `desc`: collection description
  - `urls`: a list of data source links
  - `data_keys`: the number of data keys within the collection

```{r}
pubdata::meta(collection = "ers_rural", print = TRUE)
```

For *datasets* and raw data files, `meta()` will provide:

  - `desc`: dataset description
  - `type`: "raw" or "table"
  - `path`: local file path
  - `depends`: key of the raw data file
  - `schema`: dataset variables with their descriptions and data types
  - `mask`: key naming pattern, for internal use
  - `keys`: key parts parsed with mask, for internal use

```{r}
pubdata::meta(collection = "ers_rural", key = "ruc_2023", print = TRUE)
```

`meta()` invisibly returns the information as a nested list, which can be used to programmatically extract specific components.
Set `print = FALSE` to silence the output.

```{r}
ruc_2023_meta <- pubdata::meta(collection = "ers_rural", key = "ruc_2023", print = FALSE)
ruc_2023_meta$schema$ruc_desc_2023
```



## `get()`: retrieve a dataset

`get()` for raw data keys downloads the file to local cache and returns its path.

```{r}
pubdata::get(collection = "ers_rural", key = "raw_ruc_2023")
```

For tables, on the first call `get()` downloads the source dependency file if necessary, reads it, cleans the data and stores it to a local Parquet file.
Subsequent calls load the Parquet file directly.

```{r}
pubdata::get(collection = "ers_rural", key = "ruc_2023") |>
  head(10)
```


## `path()`: dataset file path

A common usage pattern with large datasets is to query a subset from disk without loading the entire data frame into memory.
Once the dataset has been built with `get()`, the `path()` function provides a shorthand to the Parquet file location on disk.

```{r}
#| tbl-cap: Private employment by industry in Dane county, WI in 2024 (QCEW)
# slow, run once
# pubdata::get("qcew", "naics_ann_2024")

# fast
pubdata::path("qcew", "naics_ann_2024") |>
  arrow::open_dataset() |>
  filter(area_fips == "55025", agglvl_code == "74", own_code == "5") |>
  select(industry_code, annual_avg_estabs, annual_avg_emplvl) |>
  collect() |>
  left_join(pubdata::get("qcew", "meta_naics_ind"), by = "industry_code") |>
  select(sector = industry_title, establishments = annual_avg_estabs, employment = annual_avg_emplvl)
```