---
title: "QCEW: Quarterly Census of Employment and Wages"
format:
  html:
    toc: true
    toc-location: right-body
    code-fold: true
    code-overflow: wrap
    embed-resources: true
    df-print: kable
execute:
  cache: true
---

```{r}
#| echo: false
#| output: false
library(tidyverse)
library(scales)
library(arrow)
library(DT)

COLLECTION <- "qcew"
```


# Overview

> The Quarterly Census of Employment and Wages (QCEW) program publishes a quarterly count of employment and wages reported by employers covering more than 95 percent of U.S. jobs, available at the county, MSA, state and national levels by industry.

Data source links.

```{r}
#| echo: false
#| results: asis
pubdata::meta(COLLECTION, print = FALSE)$urls %>%
  iwalk(\(url, name) {
    cat(str_glue("  - [{name}]({url})"), "\n")
  })

```


# Tables

List of all tables in the collection.

```{r}
#| echo: false
#| column: page
# eval: false

descriptions <- pubdata::ls(COLLECTION, detail = TRUE) %>%
  select(key, description) %>%
  filter(!str_starts(key, "raw"))

dimensions <- descriptions$key %>%
  map(\(key) {
    pq_path <- pubdata::path(COLLECTION, key)
    if (!file.exists(pq_path)) pubdata::get(COLLECTION, key)
    df <- open_dataset(pq_path)
    list(key = key, rows = nrow(df), cols = ncol(df))
  }) %>%
  bind_rows()

inner_join(descriptions, dimensions, by = "key") %>%
  datatable()

```


# Columns

Listing of all columns in each table.

```{r}
#| echo: false
#| column: page
# eval: false

pubdata::ls(COLLECTION) %>%
  grepv("^raw_", ., invert = TRUE) %>%
  set_names() %>%
  map(\(key) {
    schema <- pubdata::meta(COLLECTION, key, print = FALSE)$schema %>%
      bind_rows(.id = "column") %>%
      relocate(column, type, desc)
    stats <- pubdata::get(COLLECTION, key) %>%
      map(\(values) list(na = sum(is.na(values)), unique = length(unique(values)))) %>%
      bind_rows(.id = "column")
    inner_join(schema, stats, by = "column")
  }) %>%
  bind_rows(.id = "key") %>%
  datatable(filter = "top")
```


# Examples

## Comparing data access methods

*By Isabella Jiang*

- Reading an full year of QCEW data using `pubdata::get()` takes a few seconds, which adds up doing multiple analysis over multiple years.
- To improve efficiency, we can combine `pubdata::path()` (returns the exact disk path for a given table table) and `arrow::open_dataset` (analyze large, multifile dataset directly from disk without reading them fully into memory)

```{r}
# Reading an entire year and filter (2020)
time_entire <- system.time({
  df_full <- pubdata::get("qcew", "naics_ann_2020") %>%
    filter(agglvl_code == "10", own_code == "5") %>%
    select(area_fips, annual_avg_emplvl, avg_annual_pay)
  })

# arrow + filter
time_arrow <- system.time({
  df_arrow <- pubdata::path("qcew", "naics_ann_2020") %>%
    arrow::open_dataset() %>%
    filter(agglvl_code == "10", own_code == "5") %>%
    select(area_fips, annual_avg_emplvl) %>%
    collect()
  })

tibble(
  Method = c("Load entire table into memory", "Query subset using Arrow Dataset"),
  Runtime = c(time_entire[["elapsed"]], time_arrow[["elapsed"]])
)
```


## Smallest states by employment in 2024

```{r}
pubdata::get("qcew", "naics_ann_2024") %>%
  filter(agglvl_code == "50") %>%
  left_join(pubdata::get("qcew", "meta_naics_area"), by = "area_fips") %>%
  select(state = area_title, employment = annual_avg_emplvl) %>%
  mutate(state = str_remove(state, " -- Statewide")) %>%
  arrange(employment) %>%
  head(10)
```


## Top 10 industries by annual average pay in 2024

*By Isabella Jiang*

```{r}
pubdata::get("qcew", "naics_ann_2024") %>% 
  filter(agglvl_code == "17") %>% 
  left_join(pubdata::get("qcew", "meta_naics_ind"), by = "industry_code") %>%
  select(industry_title = industry_title, avg_annual_pay = avg_annual_pay) %>%
  arrange(desc(avg_annual_pay)) %>% 
  head(10)
```


## Employment growth rate by ownership (2020-2024)

*By Isabella Jiang*

Using `pubdata::path()` with `arrow::open_dataset()` for faster multiyear querying.

```{r}
df_arrow <- pubdata::path("qcew", "naics_ann_2020") %>% 
  dirname() %>%
  arrow::open_dataset() %>%
  filter(
    agglvl_code == "11",
    year %in% 2020:2024,
    stringr::str_detect(own_code, "^[1-3|5]$")) %>%
  left_join(pubdata::get("qcew", "meta_naics_own"), by = "own_code") %>%
  select(year, own_title, oty_annual_avg_emplvl_pct_chg) %>%
  collect()

ggplot(df_arrow, aes(x = year, y = oty_annual_avg_emplvl_pct_chg, color = own_title)) +
  geom_line() +
  geom_point() + 
  labs(
    title = "Employment Growth Rate by Ownership (2020–2024)",
    x = "Year",
    y = "Annual Average Employment \n Percentage Change (%)",
    color = "Ownership Type"
  ) +
  theme_minimal()
```


## The annual average pay trends of top 3 supersectors in 2024

*By Isabella Jiang*

```{r}

top3_industries <- pubdata::get("qcew", "naics_ann_2024") %>%
  filter(agglvl_code == "13") %>%
  arrange(desc(avg_annual_pay)) %>%
  slice_head(n = 3) %>%
  pull(industry_code)

df_top3 <- map_dfr(2020:2024, function(y) {
  pubdata::get("qcew", paste0("naics_ann_", y)) %>%
    filter(agglvl_code == "13", industry_code %in% top3_industries) %>%
    group_by(year = y, industry_code) %>% 
    summarize(avg_annual_pay = mean(avg_annual_pay, na.rm = TRUE), .groups = "drop_last") %>% 
    left_join(pubdata::get("qcew", "meta_naics_ind"), by = "industry_code") %>%
    select(year, industry_code, industry_title, avg_annual_pay)
})

ggplot(df_top3, aes(x = year, y = avg_annual_pay, color = industry_title)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Average Annual Pay Trends for 2024's Top 3 Supersectors (2020–2024)",
    x = "Year",
    y = "Average Annual Pay",
    color = "Supersector"
  ) +
  theme_minimal()
```


## Linear association between annual average annual pay and establishments

*By Isabella Jiang*

```{r}
df_pay_estabs <- map_dfr(2020:2024, function(y) {
  pubdata::get("qcew", paste0("naics_ann_", y)) %>%
    filter(agglvl_code == "10") %>%
    mutate(year = y) %>%
    select(year, avg_annual_pay, annual_avg_estabs)
})

lm_pay_estabs <- lm(avg_annual_pay ~ annual_avg_estabs, data = df_pay_estabs)

lm_pay_estabs %>%
  summary()
```
