
---
title: "BDS Dataset Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pubdata)
```

```{r}
Sys.setenv(PUBDATA_CACHE_DIR = "/Users/divi/Desktop/APL Research/pubdata_cache")

```

```{r}
Sys.getenv("PUBDATA_CACHE_DIR")
```

```{r}
library(pubdata)
pubdata::ls("bds","raw_2022_met")
pubdata::meta("bds", "raw_2022_met")

```
Steps:
1) Look at metadata for all the datasets in bds -- gloassary website what each column means. 
#Firms
Values:
Description: A simple count of the number of firms in the cell. For state level tables, a firm with establishments
in multiple states be counted multiple times, once in each state, irrespective of the portion of the
firm residing in that state.
estabs

#Estabs
Values:
Description: A simple count of the number of establishments in the cell.
#Emp
Values:
Description: Paid employment consists of full and part-time employees, including salaried officers and
executives of corporations, who are on the payroll in the pay period including March 12. Included
are employees on paid sick leave, holidays, and vacations.
#Denom
Values:
Description: Davis-Haltiwanger-Schuh (DHS) denominator. For time t, denom is the average of employment
for times t and t-1. This variable attempts to prevent transitory shocks from creating a bias to the
relationship between net growth from t-1 to t and size.
#estabs_entry
Values:
Description: A count of establishments born within the cell during the last 12 months.
#estabs_entry_rate
Values:
Description: 100 * (estabs_entry at time t divided by the average of estabs at t and t
#estabs_exit
Values:
Description: A count of establishments exiting from within the cell during the last 12 months.
#estabs_exit_rate
Values:
Description: 100 * (estabs_exit at time t divided by the average of estabs at t and t-1).
#job_creation
Values:
Description: Count of all employment gains within the cell from expanding and opening establishments
between the week of March 12 of the prior year to the current year.
#job_creation_births
Values:
Description: Count of employment gains within the cell from establishment openings (births) between the week
of March 12 of the prior year to the current year.
#job_creation_continuers
Values:
Description: Count of employment gains within the cell from continuing establishments that expanded between
the week of March 12 of the prior year to the current year.
#job_creation_rate_births
Values:
Description: 100 * (job_creation_births / denom)
#ob_creation_rate
Values:
Description: 100 * (job_creation / denom)
#job_destruction
Values:
Description: Count of all employment losses with closing jobs. 
#job_destruction_deaths
Values:
Description: Count of all employment losses within the cell from establishment closings (deaths) between the
week of March 12 of the prior year to the current year.
#job_destruction_continuers
Values:
Description: Count of employment losses within the cell from continuing establishments that contracted
between the week of March 12 of the prior year to the current year.
#job_destruction_rate_deaths
Values:
Description: 100 * (job_destruction_deaths / denom)
#job_destruction_rate
Values:
Description: 100 * (job_destruction / denom)
#Metro: 
M=Metropolitan cities 
N= Non metropolitan areas
SW=Statewide with no location fixed. 
U=Unclassified locations. 

2) Explore the structure of 2-3 datasets and answer some questions like:
    what each column means- datatype for each column 
    trend in recent years before covid vs after covid 
    for firms/employment/ entrey rate vs exit rate of establishments. 
3) visualize the above. 
4) yaml for each of the datasets and list possible relations




```{r}
df=pubdata::get("bds","raw_2022_met")
df
```

```{r}
library(readr)
csv_url="/Users/divi/Desktop/APL Research/pubdata_cache/bds/raw/2022/bds2022_met.csv"
df <- read_csv(csv_url)
head(df)

```

```{r}
str(df$estabs_exit_rate)
```
```{r}
library(readr)
csv_url="/Users/divi/Desktop/APL Research/pubdata_cache/bds/raw/2022/bds2022_met.csv"
df <- read_csv(csv_url)
head(df)
```

# Looking at pre/post covid for establishments exit rate in metro vs non metro areas. 
Looking at years 2016-2020 vs 2020-2022.
```{r}
library(dplyr)
library(readr)
library(ggplot2)
df <- df %>%
  filter(year >= 2016 & year <= 2022) %>%
  mutate(
    covid_period = case_when(
      year < 2020 ~ "pre_covid",
      TRUE ~ "post_covid"
    )
  )
#Checking if estabs_exit_rate is numeric
str(df$estabs_exit_rate)
# it is a chr--converting to numeric. 
df <- df %>%
  mutate(estabs_exit_rate = as.numeric(estabs_exit_rate))
#cleaning: 
#df <- df %>%
  #mutate(estabs_exit_rate = parse_number(estabs_exit_rate))
df <- df %>%
  mutate(metro_status = case_when(
    metro == "M" ~ "metro",
    metro == "N" ~ "nonmetro",
    metro == "U" ~ "unclassified",
    metro == "SW"~"Statewide",
  ))

# Group by metro vs nonmetro and covid period, and summarize average exit rate
summary_df=df %>%
  group_by(year, metro) %>%
  summarise(avg_exit_rate = mean(estabs_exit_rate, na.rm = TRUE), .groups = "drop")

summary_df
# Bar plot: Exit rate by year and metro/non-metro
ggplot(summary_df, aes(x = factor(year), y = avg_exit_rate, fill = metro)) + geom_col(position = "dodge") + 
   #side-by-side bars for metro and non-metro
  geom_vline(xintercept = which(sort(unique(summary_df$year)) == 2020), linetype = "dashed", color = "red", size = 1) + labs( title = "Average Establishment Exit Rate by Year and Metro Status", x = "Year", y = "Average Exit Rate", fill = "Metro Status" ) + theme_minimal()

```
There was no signifcant change in establishment exit rates for metro vs metro pre and post covid. 
```{r}
library(readr)
csv_url="/Users/divi/Desktop/APL Research/pubdata_cache/bds/raw/2022/bds2022_met.csv"
df <- read_csv(csv_url)
head(df)
```

# Look at relationship between job creation rate and establishments exit rate in metro vs non metro areas. 
```{r}
df_processed <- df %>%
  mutate(
    # First, ensure job_creation_rate and estabs_exit_rate are numeric
    # suppressWarnings is used in case there are non-numeric values that become NA
    job_creation_rate = suppressWarnings(as.numeric(job_creation_rate)),
    estabs_exit_rate = suppressWarnings(as.numeric(estabs_exit_rate)),
    # Now classify the metro column
    metro_status = case_when(
      metro == "M" ~ "Metro Area",       # Renaming for clarity in plot legend
      metro == "N" ~ "Non-Metro Area",   # Renaming for clarity in plot legend
      metro == "U" ~ "Unclassified",
      metro == "SW" ~ "Statewide",
      TRUE ~ "Other/Unknown" # Catches any other values not explicitly handled
    )
  ) %>%
  # Remove rows where job_creation_rate or estabs_exit_rate might have become NA due to coercion
  filter(!is.na(job_creation_rate) & !is.na(estabs_exit_rate))
ggplot(df_processed, aes(x = job_creation_rate, y = estabs_exit_rate, color = metro)) +
  geom_point(alpha = 0.6, size = 2) + # Add points for each data entry, with some transparency
  geom_smooth(method = "lm", se = FALSE, aes(color = metro)) + # Add a linear trend line for each metro group
  labs(
    title = "Relationship Between Job Creation and Establishment Exit Rates",
    subtitle = "Grouped by Metro Status",
    x = "Job Creation Rate",
    y = "Establishment Exit Rate",
    color = "Metro Status" # Legend title for the color aesthetic
  ) + scale_x_continuous()
  scale_color_brewer(palette = "Set1") + # Use a colorblind-friendly palette
  theme_minimal() + # Apply a minimalist theme for a clean look
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"), # Center and bold the main title
    plot.subtitle = element_text(hjust = 0.5), # Center the subtitle
    legend.position = "bottom" # Place the legend at the bottom
  )
```
For metro cities the job creation rate and teh establishment exit rate both seem low and the slope is neither negative or positive so there is no significant relationship between job creation rate and establishment exit rate. But for unclassified cities there seems to be an outlier and hence th ehigher establishment exit rates but there is also a slightly postive relationship between job creation rate adn establishment exit rates. This could be looked into further. 

```{r}
df
```

#BDS-Business dynamics statistics 
```{r}
pubdata::meta("bds", "raw_2022_met")

```

```{r}
pubdata::meta("bds","raw_2022_st_cty")
```

```{r}
pubdata::ls("bds")
```

```{r}
install.packages("yaml")
library(yaml)

```

```{r}
meta_path <- system.file("extdata/bds/meta.yml", package = "pubdata")
meta <- yaml::read_yaml(meta_path)

```

```{r}

```

